---
title: "Model"
author: "Zhengyong Chen"
date: "2025-03-27"
output: html_document
---
## 1. Load Required Packages
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(rlang)
library(reshape2)
library(RColorBrewer)
library(mgcv)
library(caret)
```

## 2. Data Preparation
```{r}
# Load data
load("dat1.RData")

# Define response variable
response_var <- "log_antibody"

# Define continuous variables
continuous_vars <- c(
  "age",      # Age of the participant
  "height",   # Height in cm
  "weight",   # Weight in kg
  "bmi",      # Body Mass Index
  "SBP",      # Systolic Blood Pressure
  "LDL",      # Low-Density Lipoprotein
  "time"      # Time measurement
)

# Define categorical variables
categorical_vars <- setdiff(names(dat1), c("id", response_var, continuous_vars))
categorical_vars <- categorical_vars[sapply(dat1[categorical_vars], function(x) is.numeric(x) || is.factor(x))]

# Convert categorical variables to factors
dat1[categorical_vars] <- lapply(dat1[categorical_vars], factor)

# Print variable types for verification
cat("Response variable:", response_var, "\n")
cat("Continuous variables:", paste(continuous_vars, collapse = ", "), "\n")
cat("Categorical variables:", paste(categorical_vars, collapse = ", "), "\n")
```

## 3. Exploratory Data Analysis

### 3.1 Relationship between Continuous Variables and Response Variable
```{r}
# Prepare data for plotting
long_df <- dat1 %>%
  select(all_of(c(response_var, continuous_vars))) %>%
  pivot_longer(cols = all_of(continuous_vars), 
              names_to = "Variable", 
              values_to = "Value")

# Plot relationships between all continuous variables and response variable
ggplot(long_df, aes(x = Value, y = .data[[response_var]])) +
  geom_point(alpha = 0.3, color = "grey30") +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +    
  geom_smooth(method = "loess", se = FALSE, color = "blue") +                    
  facet_wrap(~ Variable, scales = "free_x", ncol = 3) +
  theme_minimal(base_size = 14) +
  labs(title = paste("Linear vs Nonlinear Relationships with", response_var),
       x = "Predictor", 
       y = paste("Log of", response_var))
```

### 3.2 Individual Variable Relationship Plots
```{r}
# Create individual relationship plots for each continuous variable
for (var in continuous_vars) {
  p <- ggplot(dat1, aes_string(x = var, y = response_var)) +
    geom_point(alpha = 0.2, color = "grey30") +
    geom_smooth(method = "lm", se = FALSE, color = "red", size = 1) +
    geom_smooth(method = "loess", se = FALSE, color = "blue", linetype = "dashed") +
    theme_bw(base_size = 16) +
    labs(title = paste(response_var, "vs", var),
         x = var,
         y = paste("Log of", response_var))
  
  print(p)
}
```

### 3.3 Correlation Analysis
```{r}
# Prepare numeric variables data
num_vars <- c(response_var, continuous_vars)
df_num <- dat1[, num_vars]

# Calculate correlation matrix
cor_matrix <- round(cor(df_num, use = "complete.obs"), 2)

# Convert to long format and create heatmap
cor_melted <- melt(cor_matrix)
ggplot(cor_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1,1), space = "Lab",
                       name="Correlation") +
  geom_text(aes(label = value), color = "black", size = 4) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(title = paste("Correlation Matrix of", response_var, "and Continuous Variables"),
       x = "", y = "")
```



# -----------
```{r}
ctrl <- trainControl(method = "cv", number = 10)
```



### GAM (mgcv)
```{r}
# Fit the GAM model with selected predictors
gam_model <- gam(log_antibody ~ age + height + s(bmi) + SBP + LDL + s(time) +
                 gender + race + smoking + diabetes + hypertension,
                 data = dat1)
```

Avoid multicollinearity:
Since BMI is a function of weight and height, so we only use height and drop weight.

### GLM
```{r}
set.seed(1)
model.glm <- train(
  log_antibody ~ age + height + bmi+ SBP + LDL + time + gender + smoking + race + diabetes + hypertension,
  data = dat1,
  method = "glm", 
  trControl = ctrl
)
```


### Glmnet
```{r}
model.glmnet <- train(
  log_antibody ~ age + height + bmi + SBP + LDL + time +
    gender + smoking + race + diabetes + hypertension,
  data = dat1,
  method = "glmnet",
  trControl = ctrl,
  tuneLength = 10  
)
```


### GAM (caret)
```{r}
set.seed(1)
model.gam <- train(
  log_antibody ~ age + height + bmi+ SBP + LDL + time + gender + smoking + race + diabetes + hypertension,
  data = dat1,
  method = "gam", 
  trControl = ctrl
)
```

### MARS
```{r}
set.seed(1)
model.mars <- train(
  log_antibody ~ age + height + bmi+ SBP + LDL + time + gender + smoking + race + diabetes + hypertension,
  data = dat1,
  method = "earth",
  tuneGrid = expand.grid(degree = 1:4, nprune = 2:20),
  trControl = ctrl
)
```

```{r}
plot(model.mars)
```

### Model Comparison
```{r}
model_list <- list(
  GLM = model.glm,
  GLMNET = model.glmnet,
  GAM = model.gam,
  MARS = model.mars
)

res <- resamples(model_list)
summary(res)
bwplot(res)
```


### Using test data

```{r}
load("dat2.RData")
```

```{r}
# Match factor levels to those in dat1
for (var in categorical_vars) {
  dat2[[var]] <- factor(dat2[[var]], levels = levels(dat1[[var]]))
}
```


```{r}
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2, na.rm = TRUE))
}

results <- data.frame(Model = character(), RMSE = numeric(), R2 = numeric())

# Loop through models
for (model_name in names(model_list)) {
  model <- model_list[[model_name]]
  preds <- predict(model, newdata = dat2)
  
  rmse_val <- rmse(dat2$log_antibody, preds)
  
  results <- rbind(results, data.frame(Model = model_name, RMSE = rmse_val))
}

print(results)
```

```{r}
for (model_name in names(model_list)) {
  dat2[[paste0("pred_", model_name)]] <- predict(model_list[[model_name]], newdata = dat2)
}

# Example plot for MARS
ggplot(dat2, aes(x = log_antibody, y = pred_MARS)) +
  geom_point(alpha = 0.3, color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "MARS: Observed vs Predicted on Test Data",
       x = "Observed log_antibody", y = "Predicted")
```


